\documentclass[english, no-theorem-numbers]{short-notes}

\usepackage{math-alg,math-ag}

\addbibresource{math.bib}
%\bibliography{math.bib}

\title{Spectral geometry of Hecke algebras and traces}
\author{David Nadler\\ Lecture notes by Clemens Koppensteiner}
\date{July 1--8, 2013}
\version{0}

\begin{document}

\newcommand\Sing{\operatorname{Sing}}
\newcommand\Loc{\operatorname{Loc}}
\newcommand\K{\sheaf K}%
\newcommand\ld[1]{\check{#1}}
\renewcommand\Gr{\operatorname{Gr}}
\newcommand\IC{\operatorname{IC}}
\newcommand\Bun{\operatorname{Bun}}

\maketitle

\tableofcontents
\bigskip\noindent
By geometry we mean that we work in the function field case, with a base field $k$ of characteristic $0$.

\section{Derived algebraic geometry}

While algebraic geometers ostensibly study solutions to polynomial equations, it has been long known that it is important to keep track of the actual equations instead of merely their solutions.
Hence we can say that Algebraic Geometry is the study of \emph{systems} of polynomial equations.
An example of our slogan \enquote{equations over solutions} is the importance of nilpotents.

General references for this section are the works of Toën--Vezzosi and Lurie.

\subsection{Three examples of \enquote{derived equations}}

\subsubsection{Intersection multiplicities}

Consider $\as 4 = \Spec k[x,y,z,w]$ and the subvarieties $X₁ = \{x = y = 0\} \cup \{z=w=0\} = \{xz=xw=yz=yw\}$ and $X₂ = \{x = z,\, y = w\}$.
Obviously, $X₁ \cap X₂ = \{0\}$.
However, if we perturb $X₂$ by a little bit, we get two points.

If we impose all equations for $X₁\cap X₂$, then we get $x² = xy = xy = y²$.
Note that we get $xy$ in two different ways: from $xw = 0$, $y = w$ and from $yz = 0$, $x = z$.
Thus the dimension of $\O_{X₁ \cap X₂}$ is $3$ instead of the expected $2$ in the perturbed, \enquote{transversal} situation.
Serre's intersection multiplicity formula tells us that 
\[ 2 = 3 - 1, \]
where the $1$ is the dimension of a $\Tor$.

\subsubsection{Base change}

Consider a Cartesian diagram
\[
    \begin{tikzpicture}
        \matrix[commutative diagram] (m) {
            X₁ ×_Y X₂ & X₂ \\
            X₁ & Y \\
        };
        \path[commutative diagram arrows]
            (m-1-1) edge node[above] {$p₂$} (m-1-2)
                    edge node[left] {$p₁$} (m-2-1)
            (m-1-2) edge node[right] {$f₂$} (m-2-2)
            (m-2-1) edge node[below] {$f₂$} (m-2-2);
    \end{tikzpicture}
\]
Base change is the statement that ${p₂}_*p₁^* = f₂^*{f₁}_*$ as functors $\catCoh{X₁} → \catCoh{X₂}$.
Here and in the rest of the document all categories of modules and all functors between them are automatically derived, even when not explicitly mentioned in the notation.

Unfortunately, base change fails even in simple situations.
\begin{Ex}
    Consider the diagram 
    \[
        \begin{tikzpicture}
            \matrix[commutative diagram] (m) {
                \{0\} ×_{\as 1} \{0\} & \{0\} \\
                \{0\} & \as 1 \\
            };
            \path[commutative diagram arrows]
            (m-1-1) edge (m-1-2)
                    edge (m-2-1)
            (m-1-2) edge (m-2-2)
            (m-2-1) edge (m-2-2);
        \end{tikzpicture}
    \]
    Classically, the upper left corner is simply equal to $\{0\}$ again.
    The functor $\catModules{k} → \catModules{k}$ around the upper left part of the diagram is simply the identity, while the functor around the lower right is given by tensoring with the complex $k[1]\oplus k$.
    Derived algebraic geometry will change the upper functor ($\id$ in this case) by changing the fiber product.
\end{Ex}

\begin{Exercise}
    Let $Q^3 = \{A ∈ M₂ : \det A = 0\}$.
    This has two canonical resolutions $Q^3_+$ and $Q³_-$ given by adding a line in the kernel or cokernel of the matrix (they are a basic example of a \enquote{flop}).
    Find the correct (derived) fiber product $\tilde Q = Q³_+ ×_{Q³} Q³_-$, so that base change holds.
    (You will need material discussed later in this section to do the exercise.)
\end{Exercise}

\subsubsection{Moduli problems}

Classically there are strange ambiguities, in particular for deformations.

\begin{Ex}
    Let $\mathcal M_n$ be the moduli space of two commuting operators on $k^n$.
    We can write $\mathcal M_n$ as a fiber product
    \[
        \begin{tikzpicture}
            \matrix[commutative diagram] (m) {
                \mathcal M_n & M_n^2 \\
                \{0\} & M_n \\
            };
            \path[commutative diagram arrows]
            (m-1-1) edge (m-1-2)
                    edge (m-2-1)
            (m-1-2) edge node[right] {$[\,{,}\,]$} (m-2-2)
            (m-2-1) edge (m-2-2);
        \end{tikzpicture}
    \]
    For $n = 1$ we obtain
    \[
        \begin{tikzpicture}
            \matrix[commutative diagram] (m) {
                \mathcal M₁ & \as 1 × \as 1 \\
                \{0\} & M₁ \\
            };
            \path[commutative diagram arrows]
            (m-1-1) edge (m-1-2)
                    edge (m-2-1)
            (m-1-2) edge node[right] {$[\,{,}\,]$} (m-2-2)
            (m-2-1) edge (m-2-2);
        \end{tikzpicture}
    \]
    and, since any two matrices in $M₁$ commute, that $\mathcal M₁ = \as 1 × \as 1$.
    But if we deform slightly and require that $[\,{,}\,] = λ \ne 0$, then the moduli space suddenly becomes empty!
\end{Ex}

\subsection{Local (affine) DAG}

Imposing equations translates to tensor products and in a derived setting this means that we have to keep track of $\Tor$s.

\begin{Def}
    A \emph{commutative differential-graded algebra} (\emph{cdga}) $A^\cx$ over $tk$ is a commutative, associative algebra in $k$-chain complexes.
\end{Def}

\begin{Rem}
    We consider cdgas only up to quasi-isomorphism, but be aware that $A^\cx$ knows more that $H^\cx(A^\cx)$.
\end{Rem}

\begin{Exercise}
    Give an example of two cdgas with the same cohomology algebra (hint: Massey products).
\end{Exercise}

\begin{Def}
    A \emph{derived ring} $A^\cx$ is a cdga which is connective, i.e.~$H^iA^\cx = 0$ for $i > 0$.
    The opposite category to derived rings is called the category of \emph{affine derived schemes}.
\end{Def}

\begin{Notation}
    We often write $π_i = H^{-i}$.
\end{Notation}

\begin{Rem}
    We have a canonical map $A^\cx → π₀A^\cx$.
    Hence we can view $A^\cx$ as a \enquote{derived thickening} of $π₀A^\cx$.
    On schemes, we see this as an embedding $\Spec π₀A^\cx \hookrightarrow \Spec A^\cx$.
\end{Rem}

The source of all examples for derived rings are iterated fiber products where we keep track of $\Tor$ complexes.

\subsubsection{Structure of cdgas}

The category of cdgas should be properly viewed as an $∞$-category and is a simplicial localization of a model category.
The main structure are simplicial sets of $\Hom$:
\[
    \Hom^n(A,B) = \Hom^0(A, B \otimes Ω^\cx(Δ^n)).
\]
Practically, to calculate derived operations we only need to know the following:
\begin{enumerate}
    \item All objects are fibrant.
    \item Cell cdgas are cofibrant.
\end{enumerate}

\begin{Def}
    A cdga $A$ is a \emph{cell cdga} if it has a filtration
    \[
        k = A_0 \hookrightarrow A₁ \hookrightarrow A₂ \hookrightarrow \dotsb \hookrightarrow A_n = A,
    \]
    such that $A_{i+1} = A_i[x_i]$ with $\deg x_i ∈ ℤ$ and $dx_i ∈ A_i$.
    \todo{Is this the same as semifree algebras?}%
    \todo{Rather than $k$ should we start with the algebra over which we are taking the tensor product?}%
\end{Def}

\begin{Ex}
    Consider a hypersurface $X = \{f = 0\}$ in $\as n$.
    Traditionally we have $\O(X) = \rquot{k[x₁,\dotsc,x_n]}{(f)}$.
    The cofibrant replacement of this is $k[x₁,\dotsc,x_n][ε]$ with $\deg ε = -1$ (and hence $ε²=0$) and $dε = f$.
\end{Ex}

\subsubsection{Linearization}

Let $A$ be a derived ring.

\paragraph{Modules}
Attached to $A$ are various categories of modules, which are summarized in Figure~\ref{fig:modules}.
\begin{figure}[htb]
    \centering
    \begin{tikzpicture}
        \node (QCoh) at (0,0) {$A\text{-}\cat{Mod}$};
        \node at (0,-0.5) [rotate=90] {$\subseteq$};
        \node (Perf) at (0,-1) {$A\text{-}\cat{Perf}$};
        \draw[->] (Perf) to[bend left=30] node[left] {\small co-completion} (QCoh);
        \draw[->] (QCoh) to[bend left=30] node[right] {\small compacts} (Perf);
        \node at (0,-1.6) {\enquote{functions}};
        \begin{scope}[xshift={5cm}]
            \node (IndCoh) at (0,0) {$\cat{IndCoh}$};
            \node at (0,-0.5) [rotate=90] {$\subseteq$};
            \node (Coh) at (0,-1) {$A\text{-}\cat{Coh}$};
            \draw[->] (Coh) to[bend left=30] node[left] {\small co-completion} (IndCoh);
            \draw[->] (IndCoh) to[bend left=30] node[right] {\small compacts} (Coh);
            \node at (0,-1.6) {\enquote{distributions}};
        \end{scope}
    \end{tikzpicture}
    \caption{Categories of modules attached to a derived ring.}
    \label{fig:modules}
\end{figure}
\begin{itemize}
    \item $A\text{-}\cat{Mod}$ is the category of derived dg modules.
    \item $A\text{-}\cat{Perf}$ the subcategory consisting of summands of finite complexes of shifts of $A$.
    \item $\cat{IndCoh}$ is the category of ind-coherent modules.
    \item $A\text{-}\cat{Coh}$ is the subcategory of $A\text{-}\cat{Mod}$ consisting of complexes with bounded coherent cohomology.
\end{itemize}

\paragraph{Cotangent complex}
To $A$ we attach its \emph{cotangent complex} $\mathbb L_A ∈ A\text{-}\cat{Mod}$, which is a connective module such that $π₀\mathbb L_A$ coincides with the cotangent bundle $Ω_{π₀A}$.
The universal characterization of the cotangent complex is 
\[
    \operatorname{Der}(A,M) = \Hom(\mathbb L_A, M).    
\]
In particular we have a \enquote{Kähler differential} $d\colon A → \mathbb L_A$.

To calculate we only need to know that $\mathbb L_A$ coincides with the cotangent bundle on smooth classical affine schemes and if $D = A \otimes_C^{\mathbb L} B$, then
\[
    \mathbb L_D = \operatorname{Cone}\left( \mathbb L_A \otimes_A D → (\mathbb L_B \otimes_B D) \oplus (\mathbb L_C \otimes_C D) \right),
\]
where the map is given by pullback.

\subsection{Global DAG}

We take the functor of points viewpoint:
\[
    \begin{tikzpicture}
        \matrix[commutative diagram] (m) {
            \cat{Aff}^{\mathrm{op}} & \cat{Sets} \\
            \cat{DAff}^{\mathrm{op}} & \cat{Spaces} = \cat{SSet} \\
        };

        \path[->] 
            (m-1-1) edge node[above] {Sch} (m-1-2)
                    edge node[auto,sloped,anchor={south}] {Stacks} (m-2-2)
            (m-2-1) edge node[below] {DStacks} (m-2-2);
        \path[right hook->]
            (m-1-1) edge (m-2-1)
            (m-1-2) edge (m-2-2);
    \end{tikzpicture}
\]
In particular, Artin derived stack are given by a diagram of the form
\[
    X \leftarrow \left(X₀ \leftleftarrows X₁\right),
\]
where $X₀$ and $X₁$ are derived schemes and the two arrows $\leftleftarrows$ are smooth maps.

\begin{Ex}\leavevmode
    \begin{enumerate}
        \item $BG \leftarrow (\mathrm{pt} \leftleftarrows G$), for $G$ a group.
        \item $M_{1,1} \leftarrow (? \leftleftarrows ?)$.
        \item $S¹ \leftarrow (\mathrm{pt} \leftleftarrows ℤ)$.
            \qedhere
    \end{enumerate}
\end{Ex}

\section{Microlocal algebraic geometry}

The goal of this section is to introduce geometric Arthur parameters. 
There is work by many people on the ideas we will introduce, but the \enquote{state of the art} is \cite{ArinkinGaitsgory:arXiv:SingularSupport}.

We will work locally and set $X = \{F = 0\}$, where $F\colon \as n → \as k$.
We want to understand the (derived) category $\catCoh X$.

\subsection{Philosophy: two themes}

The ideas we will introduced a based on two major themes in mathematics. 
The first of these is illustrated by the following example.

\begin{Ex}
    Let $A$ be an algebra and consider the category $\catModules A$ of $A$-modules.
    Let $Z(A)$ be the center of $A$.
    Then $\Spec Z(A)$ is a scheme, and we can localize $A$ as a sheaf of algebras on $\Spec Z(A)$.
    Therefore we can localize the category $\catModules A$ on $\Spec Z(A)$.
\end{Ex}

More generally, if $\cat C$ is a category, we can sheafify $\cat C$ over $Z(\cat C) = \End(\id[\cat{C}])$.

The second theme is that there are two types of \enquote{geometric measurements}: 
\begin{itemize}
    \item static measurements corresponding to points of $X$, e.g.~talking stalks;
    \item dynamic measurements corresponding to $T^*X$, e.g.~taking vanishing cycles.
\end{itemize}

A classical example is Fourier transform $L²(S¹) \isoto L²(ℤ)$.

\subsection{Koszul duality}

Let $V = \Spec k[x₁,\dotsc,x_n]$ with $\deg x_i = -1$ be a \enquote{vector space in degree $-1$.}
We want to analyze $\catCoh V$.

Note that $V_{\mathrm{cl}} = \Spec π₀V$ is just a point, so that this doesn't give us any way to localize.
Thus we introduce \emph{codirections} $V^*[1] = \Spec k[ξ₁,\dotsc,ξ_n]$ with $\deg ξ_i = 2$.

\begin{Thm}[Koszul Duality]
    There is an equivalence of categories
    \[
        \catCoh V \cong \catCoh{V^*[1]},
    \]
    sending the structure sheaf $\O_V$ to the skyscraper $k$ on $V^*[1]$ and the skyscraper $k$ on $V$ to $\O_{V^*[1]}$.
\end{Thm}

Thus we can localize coherent sheaves on $V$ on $V^*[1]$ (which is given by a \enquote{usual} polynomial ring).
Supports will be closed conic subvarieties.

\begin{Exercise}
    Under Koszul Duality, $\sheaf F ∈ \catCoh V$ is perfect if and only if $\supp F \subseteq \{0\} \subseteq V^*[1]$.
\end{Exercise}

\begin{Rem}
    We think of $\catCoh V$ as \enquote{distributions} and $\catPerf V \subseteq \catCoh V$ as \enquote{functions}.
    Koszul Duality measures the codirections on which \enquote{smoothness} fails.
\end{Rem}

\begin{Exercise}
    Show that $V = \{0\} ×_{\as n} \{0\}$ as derived schemes.
    Thus $V$ is an Abelian groupoid via \enquote{concatenation of paths}.
    This turns $\catCoh V$ into a tensor category with tensor product $*$ given by convolution.
    Show that Koszul Duality is an equivalence of tensor categories exchanging $*$ with the usual tensor product $\otimes$ in $\catCoh{V^*[1]}$.
\end{Exercise}

\subsection{The scheme of singularities}

Let us return to the situation of $X = \{F = 0\} \subseteq \as n$.
The cotangent complex of $X$ is given by
\[
    \mathbb L_X = \left( Ω_{\as k} \xrightarrow{dF^*} Ω_{\as n} \right)
\]
in degrees $-1$ and $0$.

\begin{Def}
    The (classical) scheme $\Sing(X) → X$ is the scheme with fiber at $x ∈ X$ given by $H^{-1}\mathbb L_X$.
\end{Def}

\begin{Ex}
    Consider $X = \{x³ = y²\}$.
    Then $\Sing(X)$ is the whole picture given in Figure~\ref{fig:SingEx}.
    \begin{figure}[htb]
        \centering
        \begin{tikzpicture}
            \draw plot[smooth] file {cusp.table};
            \draw[y=-1cm] plot[smooth] file {cusp.table};
            \draw (-1,0) -- (1.5,0);
        \end{tikzpicture}
        \caption{Scheme of singularities of $x³ = y²$}
        \label{fig:SingEx}
    \end{figure}
\end{Ex}

\begin{Rem}
    We always have $\Sing X \subseteq X × \as k$.
\end{Rem}

We can then localize $\catCoh X$ over $\Sing(X)$.

\begin{Rem}
    This will be the ultimate localization.
\end{Rem}

To do so, we need do find the action of $\O(\as k) = k[ξ₁,\dotsc,ξ_k]$.
Consider the Cartesian diagrams (in derived schemes)
\[
    \begin{tikzpicture}
        \matrix[commutative diagram] (m) {
            X & \as n \\
            \{0\} & \as k \\
        };
        \path[commutative diagram arrows]
            (m-1-1) edge (m-1-2) edge (m-2-1)
            (m-1-2) edge node[right] {$F$} (m-2-2)
            (m-2-1) edge (m-2-2);

        \begin{scope}[xshift={4cm}]
            \matrix[commutative diagram] (m) {
                V & \{0\} \\
                \{0\} & \as k \\
            };
            \path[commutative diagram arrows]
                (m-1-1) edge (m-1-2) edge (m-2-1)
                (m-1-2) edge (m-2-2)
                (m-2-1) edge (m-2-2);
        \end{scope}
    \end{tikzpicture}
\]
We should think of points of $X$ as paths from $F$ to $\{0\}$ and points in $V$ as loops based at $\{0\}$.
Thus we can concatenate such a path with a loop and get another path.

The upshot of this is that we have an action of $\catCoh V$ on $\catCoh X$.
The unit of the convolution product $*$ of $\catCoh V$ is $k$.
Hence we obtain a map
\[
    k[ξ₁,\dotsc,ξ_k] = \End(k) → \End(\id[\catCoh X]).
\]
In other words, $k[ξ₁,\dotsc,ξ_k]$ acts universally on every object of $\catCoh X$.
Hence $\catCoh X$ can be localized on $X × \as n$.

\begin{Exercise}
    The support of any element of $\catCoh X$ will always be contained in $\Sing(X)$.
\end{Exercise}

\begin{Ex}
    Let $Q³ = \{A ∈ M₂ : \det A = 0\}$.
    This can be presented as
    \[
    \begin{tikzpicture}
        \matrix[commutative diagram] (m) {
            Q³ & \as 4 \\
            \{0\} & \as 1 \\
        };
        \path[commutative diagram arrows]
            (m-1-1) edge (m-1-2) edge (m-2-1)
            (m-1-2) edge node[right] {$\det$} (m-2-2)
            (m-2-1) edge (m-2-2);
    \end{tikzpicture}
    \]
    and $\Sing(X) \subseteq X × \as 1$ is $X$ together with a one-dimensional \enquote{spoke} at the origin.
\end{Ex}

\begin{Exercise}
    Let $\sheaf F ∈ \catCoh X$.
    Show that $\supp \sheaf F \subseteq \{0\} \subseteq \Sing(X)$ if and only if $\sheaf F ∈ \catPerf X$.
\end{Exercise}

\subsection{Geometric Arthur parameters}

In the function field setup, the analogue of the number field in the Langlands program is a smooth projective curve $C$ over the ground field $k$.
The analogue of a Galois representation is then a local system for $G$ (or rather $\ld G$, but we will only talk about one side of the correspondence for now, so we omit the dual for simplicity).
We write $\Loc_G(C)$ for the moduli of local $G$-systems on $C$.

A common source of confusion here is that there are two notions of local systems in the geometric context:
\begin{itemize}
    \item The \enquote{Betti version}: A local system is a representation of the fundamental group of $C$, i.e. a map $π₁(C) → G$.
        This looks most immediately like a Galois representation: the fundamental group is the geometric analogue of the Galois group in the number field setting.
    \item The \enquote{de Rham version}: A local system is a $G$-bundle with a flat connection.
\end{itemize}
These two notions are analytically equivalent, but not algebraically equivalent.
So their local geometry is the same, but there are global aspects where they differ.

For simplicity, we will only consider the Betti version in this lecture.
Thus we set
\[
    \Loc_G(C) = \rquot{\left\{ π₁C → G  \right\}}{G}.
\]
This is a nice Artin stack.
Let us describe it in more detail.
First, if we pick a point in $C$ and trivialize a $G$-bundle we get rid of the symmetries.
Hence we get a presentation of $\Loc_G(C)$ as the groupoid\todo{So $\Loc_G(C)'$ is the moduli of local systems trivialized at a point of $C$?}
\[
    \Loc_G(C) \leftarrow \left( \Loc_G(C)' \leftleftarrows G \right).
\]
Let $g$ be the genus of $C$ (so $2g$ is the number of generators of $π₁(C)$).
We can further write $\Loc_G(C)'$ as the fiber product
\[
    \begin{tikzpicture}
        \matrix[commutative diagram] (m) {
            \Loc_G(C)' & G^{2g} \\
            \{1\} & G \\
        };
        \path[commutative diagram arrows]
            (m-1-1) edge (m-1-2) edge (m-2-1)
            (m-1-2) edge node[right] {$[\,{,}\,]$} (m-2-2)
            (m-2-1) edge (m-2-2);
    \end{tikzpicture}
\]
where $[\,{,}\,]$ is the commutator map\todo{How to define the commutator map $G^{2g} → G$?}. 
In other words, we take the curve, puncture it, take a local system on the punctured curve which will have $2g$ monodromies and ask that it extends over the puncture, i.e. the commutator is $1$.

Now we have a diagram as before (with $F = [\,{,}\,]$).
The spaces $G$ and $G^{2g}$ are not affine, but they are still nice smooth varieties.
Hence, if we ignore the $G$-symmetries, $\Loc_G(C)$ looks like the kind of complete intersections we studied in the first part of this section.

We can now define the geometric analogue of Arthur's parameters: they are $\Sing(X)$ for this stack, i.e. we call
\[
    \Sing(\Loc_G(C)) \subseteq \rquot{(\Loc_G(C)' × \liealg g^*)}{G}
\]
the \emph{moduli of geometric Arthur's parameters}.
Note that $\liealg g^*$ appears for the cotangent space (as $G$ is not affine).
One expects that automorphic objects lead to coherent sheaves with nilpotent Arthur's parameters.
\todo{Interpretation}

\section{Geometry of Hecke categories -- Satake correspondence}

The goal of this section is discuss a geometric verison of the Satake correspondence.
A lot of the material we will talk about has been very well understood for the last two decades, but by the end we will talk about something new that will tie in with the last two sections.
The goal is to give an introduction to geometric techniques for understanding Hecke categories -- specifically spherical Hecke categories.

\subsection{Reminder on the Satake correpondence}

Let $G$ be a reductive group.
As we will soon move to characteristic $0$, so for simplicity we assume that $G$ is defined over $ℤ$.
Attached to $G$ is its root datum
\[
    \left( Λ_G,\, \ld{Λ}_G,\, R_G, \ld R_G \right),
\]
consisting of the coweight lattice (homomorphisms from $\mathbb G_m$ into a maximal torus of $G$), the weight lattice (the dual), the coroots and the roots of $G$.
Switching weights and coweights and roots and coroots, we obtain the \emph{Langlands dual group} $\ld G$ of $G$.
\begin{Ex}
    The Langlands dual of $\GL n$ is again $\GL n$.
    The Langlands dual of $\SL n$ is $\PGL n$.
\end{Ex}

Recall that the \emph{spherical Hecke algebra} of $G$ consists of the compactly supported functions on $G(ℚ_p)$ which are $G(ℤ_p)$-biinvariant, i.e.
\[
    H_{\mathrm{sph}} = ℂ_c\biggl[ \dquot{G(ℤ_p)}{G(ℚ_p)}{G(ℤ_p)} \biggr].
\]
The \emph{Satake correspondence} identifies this with functions on the Langlands dual group.
More concretely it is an isomorphism 
\begin{equation}
    \label{eq:Hecke:Satake}%
    H_{\mathrm{sph}} \cong \O(\ld G)^{\ld G}
\end{equation}
of commutative algebras between the spherical Hecke algebra of $G$ (with convolution $*$; we won't discuss the definition here, but see the geometric version below) and the representation ring of $\ld G$ (with multiplication $\cdot$).
One can think of this as a kind of Fourier transform: it identifies something that is not evidently commutative with functions on something.

The goal of this section is to discuss a geometric version of the Satake isomorphism: 
We are going replace the algebras be symmetric monoidal categories of sheaves.
Thus we get a statement with a lot more structure, that allows one to use the tools from algebraic geometry (e.g.~Hodge theory).
So on the one hand the geometric version will be a more complicated statement than \eqref{eq:Hecke:Satake}, but on the other hand its additional structure gives a richer, more powerful statement.
In particular, \eqref{eq:Hecke:Satake} can then be deduced from the categorical version.

\subsection{The affine Grassmannian}

Let $k = ℂ$ (though the construction also works for other fields).

We set $\K = k\lParen t \rParen$ be the ring of Laurent series over $k$ and $\O = k\lBrack t \rBrack$ its subring of power series.
We think of $\Spec \O$ as a formal disc and $\Spec \K$ as a punctured formal disc.
Thus we call $LG = G(\K)$ the \emph{loop group} and $L_+G = G(\O)$ the \emph{arc group}.
For example, if $G = \GL n$, then $G(\K)$ consists of invertible matrices with coefficients in $\K$.

\begin{Def}
    The \emph{affine Grassmannian} (or \emph{loop Grassmannian}) is
    \[
        \Gr = \Gr_G = \rquot{LG}{L_+G}.
    \]
\end{Def}

The affine Grassmannian is an increasing union of projective varieties.

\begin{Ex}
    Let $G = \GL n$.
    Then we have an action of $L\GL n$ on $\K^n$.
    The basic lattice $\O^n \subseteq \K^n$ (see Figure~\ref{fig:GLnlattice}) is preserved by $L_+\GL n$.
    \begin{figure}[htb]
        \centering
        \begin{tikzpicture}[y={-0.5cm}, x={0.5cm}]
            \foreach \i in {-1,0,1,2} {
                \foreach \j in {-1,0,1} {
                    \fill ({\j},{\i}) circle[radius=2pt];
                }
                \node[anchor=mid west] at (2,\i) {$t^{\i}$};
            }
            \draw[rounded corners=2pt] (-1.5,3) -- (-1.5,-0.5) -- (1.5,-0.5) -- (1.5,3);
            \draw[loosely dotted, very thick] (0,-1.5) -- (0,-2.5);
            \draw[loosely dotted, very thick] (0,2.5) -- (0,3.5);
        \end{tikzpicture}
        \caption{The lattice of basis vectors of $\K^3$ after fixing a basis for $ℂ³$ with the basic lattice $\O^n$ marked.}
        \label{fig:GLnlattice}
    \end{figure}
    \begin{Exercise}
        Show that 
        \[
            \Gr_{\GL n} = 
            \left\{ 
                \text{$k$-subspaces $W \subseteq \K^n$ such that $tW \subseteq W$ and $t^N\O^n \subseteq W \subseteq t^{-N}\O^n$ for $N \gg 0$}
            \right\}.
        \]
    \end{Exercise}
    Hence we can write $\Gr_{\GL n} = \smash{\bigcup\limits_{\mathclap{N = 0}}^∞ \Gr_N}$, where
    \[
        \Gr_N = 
        \left\{ 
            \text{$k$-subspaces $W \subseteq \K^n$ such that $tW \subseteq W$ and $t^N\O^n \subseteq W \subseteq t^{-N}\O^n$}
        \right\}.
    \]
    is a projective variety.
    \begin{figure}[htb]
        \centering
        \begin{tikzpicture}[y={-0.5cm}, x={0.5cm}]
            \foreach \i in {-2,-1,0,1,2} {
                \foreach \j in {-1,0,1} {
                    \fill ({\j},{\i}) circle[radius=2pt];
                }
                \node[anchor=mid west] at (2,\i) {$t^{\i}$};
            }
            \draw[rounded corners=2pt] (-1.6,2.5) -- (-1.6,-2.3) -- (1.6,-2.3) -- (1.6,2.5);
            \draw[rounded corners=2pt] (-1.4,2.5) -- (-1.4,1.5) -- (1.4,1.5) -- (1.4,2.5);
            \draw[loosely dotted, very thick] (0,-2.5) -- (0,-3.5);
            \draw[loosely dotted, very thick] (0,2.5) -- (0,3.5);
        \end{tikzpicture}
        \caption{The bounds for $W$ in $\Gr_2$.}
        \label{fig:GLnlatticeN}
    \end{figure}
\end{Ex}

\subsection{Alternative descriptions}

\subsubsection{Bundles}

Let $D = \Spec \O$ and $D^× = \Spec \K$.
Then $\Gr$ is the set of $G$-bundles on $D$, trivialized on $D^×$.
Hence, if $\mathbb D = D \amalg_{D^×} D$ is the disc with two origins, we have
\[
    \lquot{L_+G}{\Gr} =
    \dquot{L_+G}{LG}{L_+G} = 
    \{ \text{$G$-bundles on $\mathbb D$}\}.
\]

\subsubsection{Topological}

Let $K \subseteq G$ be a maximal compact and let $ΩK$ be the based polynomial\footnote{I.e., as maps $S¹ → K$ the loops have finite Fourier expansion} loop group on $K$.
Then
\[ \Gr \cong ΩK\]
as topological spaces.
In fact, $LG \cong ΩK × L_+G$.

\begin{Exercise}
    Show that for $G = \PGL 2$ one has
    \[
        \Gr_{\PGL 2} = \rquot{\operatorname{Free}(S²)}{xa(x) = 1},
    \]
    where $\operatorname{Free}(X)$ denotes the free topological group on $X$ and $a\colon S² → S²$ is the antipodal map.
\end{Exercise}

\subsection{A little structure theory}

\subsubsection{Cartan}

There is an isomorphism of sets
\[
    \lquot{L_+G}{\Gr} =
    \dquot{L_+G}{LG}{L_+G} \cong
    \rquot{Λ_G}{W} \cong
    Λ_G^+,
\]
where $W$ is the Weyl group and $Λ^+_G$ the dominant coweights.

\subsubsection{Iwasawa}

\[
    \lquot{N(\K)}{\Gr} = \dquot{N(\K)}{LG}{L_+G} \cong Λ_G.
\]

\todo{Example of $\PGL 2$.}

\subsection{Geometric Satake}

\subsubsection{Geometric \texorpdfstring{$L_+G$}{L+G}-invariant \enquote{distributions} on \texorpdfstring{$\Gr$}{Gr}}

These lead to the category $\cat P_{L_+G}(\Gr)$ of $L_+G$-equivariant perverse sheaves on $\Gr$.
This category is Abelian and semisimple.
The simple objects are given by the intersection complexes $\IC^λ = \IC(\Gr^λ)$, where $\Gr^λ \subseteq \Gr$ is the closure of the $L_+G$-orbit indexed by $λ ∈ \rquot{Λ_G}W$.
This is the \enquote{best basis of the Hecke algebra.}

The first hint of Satake are the following equations due to Lustzig.
Let $V^λ$ be the irreducible representation of $\ld G$ of highest weight $λ ∈ \rquot{\ld Λ_{\ld G}}W = \rquot{Λ_G}W$.
Then,
\[ 
    \dim \mathrm RΓ(\Gr,\, \IC^λ) = \dim V^λ
\]
and
\[
    \dim \res{\IC^λ}μ = \dim \operatorname{wt}_μ(V^λ)
\]
for any point $λ$ in $\Gr^μ$.

\subsubsection{Geometric Satake}

\begin{Thm}
    There is an isomorphism of tensor categories
    \[
        \cat P_{L_+G}(\Gr) \cong \cat{Rep}(\ld G).
    \]
\end{Thm}

\subsubsection{Where is the tensor structure?}

The category $\cat P_{L_+G}(\Gr)$ has a tensor structure $*$ given by the usual convolution pattern via $\lquot{L_+G}{\Gr} = \{\text{$G$-bundles on $\mathbb D$}\}$.
However there is an alternative, \enquote{better} point of view given by the \emph{fusion} product.

Recall that $\Gr$ parametrizes $G$-bundles on $D$ with a trivialization on $D^×$.
Let $C$ be and smooth curve and $c ∈ C$ a closed point.
By extending the bundles one sees that $\Gr$ is also given by $G$-bundles on $C$ with a trivialization of $C \setminus c$.
Note that from this perspective all points of $C$ look similar.

We now introduce the \emph{Beilinson--Drinfeld} (or \emph{BD}) Grassmannian.
Let $S = \{c_i : 1 \le i \le k\}$ be a collection of closed points in $C$.
Then set
\[
    \Gr^{(k)} = \left\{ \text{$G$-bundles on $C$, trivialized away from $S$} \right\}.
\]
\begin{Exercise}
    \[
        \Gr^{(k)} = \smash[t]{\prod\limits_{\mathclap{i=1}}^k \Gr}.
    \]
\end{Exercise}
Now we allow $S$ to move and points to collide. 
This yields an evidently commutative tensor product of $\cat P_{L_+G}(\Gr)$ via taking nearby cycles as points collide.

\subsubsection{Derived version}

Instead of just the Abelian category of perverse sheaves, consider the full $L_+G$-equivariant, constructible derived category $D_{\mathrm{c},\, L_+G}(\Gr)$.
We would like a spectral interpretation of this category.

In view of the $\lquot{L_+G}{\Gr} = \{ \text{$G$-bundles on $\mathbb D$}\}$, we expect 
\[
    \Loc_{\ld G}(\mathbb D) =
    \rquot{\left( \{1\} \mathbin{×\limits_{\ld G}} \{1\} \right)}{\ld G} =
    \rquot{\ld{\liealg g}[1]}{\ld G}
\]
to appear on the spectral side, where $\ld{\liealg g}[1] = \Spec\operatorname{Sym}\left( (\ld{\liealg g})^*[1] \right)$.

Consider $\catCoh{\Loc_{\ld G}(\mathbb D)}$.
Note that $H^{-1}\mathbb L_{\Loc_{\ld G}(\mathbb D)} \cong (\ld{\liealg g})^*$.
Let $N \subseteq (\ld{\liealg g})^*$ be the nilpotent cone.
\begin{Thm}
    \[
        D_{\mathrm c}(\Bun_G(\mathbb D)) \cong
        D_{\mathrm c,\, L_+G}(\Gr) \cong
        \cat{Coh}_N(\Loc_{\ld G}(\mathbb D)).
        \qedhere
    \]
\end{Thm}
\section{Geometric traces}

\printbibliography
\end{document}
